\chapter{Theory}\label{theory}

\section{Interior ballistics}\label{interior-ballistics}

\subsection{Control volume notation}\label{notation}

\input{control-volume-figure.tex}

Internally, BlasterSim simulate blasters using control volumes with arbitrary flow connections between control volumes.
This allows BlasterSim to simulates spring and pneumatic blasters with the same core simulation code. This also allows simulating more atypical blasters without major code changes.

An isolated control volume numbered $i$ is shown in \figref{control volume}.
A single flow connection to the control volume numbered $j$ is shown with
mass flow rate $\dot{m}_{i \rightarrow j}$,
a projectile/plunger labeled p with mass $m_{\text{p},i}$,
a spring with stiffness $k_i$, mass $m_{\text{s},i}$, and spring precompression $\Delta_{\text{pre},i}$ (see \figref{springer unprimed} for a graphical definition),
cross-sectional area $A_i$,
projectile/plunger position $x_i$ and velocity $\dot{x}_i$, and
gas mass $m_i$ and energy $E_i$.

The integers $i$ and $j$ will be used as indices for control volumes.

Gas mass and energy as chosen as these are conserved variables and it is believed that using conserved variables will lead to better conservation properties in the numerical methods.

A control volume represented like this can be used to represent a variety of scenarios relevant to gas gun interior ballistics, including:
\begin{itemize}
\item a barrel, where p represents the projectile, $x_i$ represents the projectile location, and $k_i$ and $m_{\text{s},i}$ are set to zero;
\item a plunger tube for springers where p represents the plunger head, and $x_i$ represents the plunger head location;
\item a constant volume chamber for pneumatics where p represents a solid wall, $m_{\text{p},i}$ is set to infinity, and $k_i$ and $m_{\text{s},i}$ are set to zero;
\item a piston tube used in HPA Nerf blasters similar to a plunger tube; and % chktex 20
\item a constant pressure atmosphere where $x_i$ is arbitrary, and $k_i$ and $m_{\text{s},i}$ are set to zero.
\end{itemize}

The gas mass could be composed of multiple species in BlasterSim.
At the moment, the pneumatic and springer modes of BlasterSim use only dry air, but humidity and alternative gases are intended to be added in the future.
For completeness, this guide will be written as if multiple gas species are present.
The mass fraction $y$ and gas species index $k$ which identifies the gas are used.

In the typeset math in this documentation, when there are separate intensive and extensive versions of the same quantity, the two will be distinguished by making the extensive version uppercase and the intensive version lowercase.
For example, internal energy per unit mass is $u$ and internal energy is $U$.
Note that this convention is not followed in the typeset math when it is conventional to make an extensive quantity lowercase (like mass $m$) or an intensive quantity uppercase (like $R$), only when there are separate intensive and extensive versions of the same quantity.
This capitalization convention is not followed in BlasterSim's source code.
See \secref{conventions} for a discussion of the capitalization convention followed in the source code.
Because each variable in BlasterSim's source code is given a unit (see \secref{unit-checking}), whether a quantity is extensive or intensive can be determined from the units for that quantity, if unclear.

\subsection{Conservation laws}\label{conservation-laws}
% TODO
% mass conservation
% energy conservation

\begin{equation}
    \dv{\dot{m}_{k,i}}{t} = \sum_j \left(y_{k,j} \dot{m}_{j \rightarrow i} - y_{k,i} \dot{m}_{i \rightarrow j}\right)\label{eq:mass conservation}
\end{equation}

\begin{equation}
    \dv{E_i}{t} = -p_i \, A_i \, \dot{x}_i + \sum_j \left(\dot{m}_{j \rightarrow i} h_j - \dot{m}_{i \rightarrow j} h_i\right)\label{eq:energy conservation}
\end{equation}

% TODO: note that definition of $E_i$ could include more than thermal energy and might in the future
\begin{equation}
    E_i = m_i \, u_i
\end{equation}

% TODO: where the pressure work term comes from, why its sign is negative; see moran_fundamentals_2008 p. 39

\subsection{Equations of state}\label{equations-of-state}
% TODO

\begin{equation}
    p_i = \frac{m_i \, R_i \, T_i}{A_i \, x_i}\label{eq:ideal gas eos}
\end{equation}

\subsection{Thermodynamic properties}\label{thermo}
% TODO

% TODO: R from \gamma
% TODO: c_v from R
% TODO: c_p from R
% TODO: mixture R, u, and h

\begin{equation}
    u_{k,i}(T) = u_{0,k} + c_{\text{v},k} \, (T_i - T_{0,k})\label{eq:internal energy}
\end{equation}

\begin{equation}
    h_{k,i}(T) = h_{0,k} + c_{\text{p},k} \, (T_i - T_{0,k})\label{eq:enthalpy}
\end{equation}

\subsection{Connection flow model}\label{connection-flow-model}
% TODO

\citet[ch.~5]{beater_pneumatic_2007}

% TODO: inlet condition, throttling process, ideal gas implies T_in = T_out

\subsection{Valve opening model}\label{valve-opening-model}
% TODO

Valves do not open instantaneously, and sometimes slow opening has a significant impact on performance.
BlasterSim models how far a valve is open with a prescribed valve opening model.
$\alpha_{i \rightarrow j}$ is the valve opening fraction, which is how open the valve is.
$\alpha_{i \rightarrow j} = 0$ is fully closed $\dot{m}_{i \rightarrow j} = 0$ and $\alpha_{i \rightarrow j} = 1$ is fully open.
The specific equation BlasterSim uses is
\begin{align}
    \alpha_{i \rightarrow j} = \alpha_{0,i \rightarrow j}
    &+ \dot{\alpha}_{0,i \rightarrow j} \, \left(\frac{t}{t_{\text{opening},i \rightarrow j}}\right) \nonumber \\
    &+ (3 - 3 \alpha_{0,i \rightarrow j} - 2 \dot{\alpha}_{0,i \rightarrow j}) \, {\left(\frac{t}{t_{\text{opening},i \rightarrow j}}\right)}^2 \nonumber \\
    &- (2 - 2 \alpha_{0,i \rightarrow j} - \dot{\alpha}_{0,i \rightarrow j}) \, {\left(\frac{t}{t_{\text{opening},i \rightarrow j}}\right)}^3,
\end{align}
for $t < t_{\text{opening},i \rightarrow j}$. For $t \geq t_{\text{opening},i \rightarrow j}$, $\alpha_{i \rightarrow j} = 1$.
$\alpha_{0,i \rightarrow j}$ is the initial (time zero) valve opening fraction (useful when a valve isn't used like in a springer).
$\dot{\alpha}_{0,i \rightarrow j}$ is the initial valve opening rate.
$t_{\text{opening},i \rightarrow j}$ is the valve opening time, the time it takes for the valve to fully open (reach $\alpha_{i \rightarrow j} = 1$).

This equation may seem overly complicated, but is the simplest polynomial equation that satisfies a few constraints:
\begin{align}
    \alpha_{i \rightarrow j}(0)                                            &= \alpha_{0,i \rightarrow j} \qquad \text{(set initial valve opening fraction)} \\
    {\dv{\alpha_{i \rightarrow j}}{t}}(0)                                  &= \dot{\alpha}_{0,i \rightarrow j} \qquad \text{(set initial valve opening rate)} \\
    \alpha_{i \rightarrow j}(t_{\text{opening},i \rightarrow j})           &= 1 \qquad \text{(valve is fully open at}~t_{\text{opening},i \rightarrow j}~\text{)} \\ % chktex 1 chktex 9
    {\dv{\alpha_{i \rightarrow j}}{t}}(t_{\text{opening},i \rightarrow j}) &= 0 \qquad \text{(needed for automatic differentiation)}
\end{align}

When the valve is fully open, $\alpha_{i \rightarrow j}$ no longer changes with time, so its derivative is zero.
Consequently, the last constraint listed is needed to match the derivative at $t = t_{\text{opening},i \rightarrow j}$, which is necessary for automatic differentiation.
See \secref{autodiff} for more about automatic differentiation in BlasterSim.
The last constraint prevents a simple linear model ($\alpha_{i \rightarrow j} = \alpha_{0,i \rightarrow j} + (1 - \alpha_{0,i \rightarrow j})\,t/t_{\text{opening},i \rightarrow j}$) from being used.

Note that to ensure monotonicity of $\alpha_{i \rightarrow j}$ (in other words, avoid oscillations of the valve opening fraction), $\dot{\alpha}_{0,i \rightarrow j}$ must satisfy the inequality $0 \leq \dot{\alpha}_{0,i \rightarrow j} \leq 3 (1 - \alpha_{0,i \rightarrow j})$. % TODO: Double check this

The pneumatic and springer cases will now be discussed, with the $i \rightarrow j$ subscript dropped for simplicity as there is only one flow restriction in both cases.

For pneumatics, $\alpha_0 = 0$ and $\dot{\alpha}_0 = 1$.
The second condition approximates a simple linear model.
In the future $\dot{\alpha}_0$ may become an input parameter if deemed necessary to improve accuracy.
The SpudFiles~Wiki~\cite{noauthor_opening_2008} gives some estimated opening times:
\begin{itemize}
\item Burst disks: likely under 1~ms
\item Pilot-operated valves (like QEVs, ``back-pressure tanks'', ``cores''): 3--5~ms
\item Ball valves: about 100~ms if hand activated
\end{itemize}
These values are recommended as starting points only.
For modeling any particular blaster, it is better to try to independently determine the valve opening time through something like high speed video.

For springers, $\alpha_0 = 1$ and $\dot{\alpha}_0 = 0$.
As there is no valve in a springer, this simply sets the flow restriction to always be open.

This simple valve opening model is most accurate for manually operated valves.
More detailed modeling of pilot-operated valves could make determining the valve opening time unnecessary, but this alternative valve opening model has not yet been added to BlasterSim.

\subsection{Projectile and plunger equations of motion}\label{equations-of-motion}
% TODO

% TODO: effective pressure of friction, $p_{\text{f},i$

\begin{align}
    \dv{x_i}{t}       &= \dot{x}_i \\
    \dv{\dot{x}_i}{t} &=
    \frac{A_i}{m_{\text{eff},i}} \left(p_i - p_{\text{mirror},i} - p_{\text{f},i}\right)
    - \frac{k_i}{m_{\text{eff},i}} \left(x_i + \Delta_{\text{pre},i}\right) \label{eq:Newton's second law}
\end{align}

The effective mass of the projectile/plunger factors in the spring mass.
The spring is not moving at a uniform velocity as one end is stationary, so it would be incorrect to add all the spring mass to the effective mass.
The effective mass equation used is
\begin{equation}
    m_{\text{eff},i} = m_{\text{p},i} + C_\text{ms} m_{\text{s},i}\label{eq:effective mass}
\end{equation}
where $C_\text{ms} = \tfrac{1}{3}$ as suggested by \citet{ruby_equivalent_2000} for a stiff spring.

\subsection{Projectile and plunger friction model}\label{friction}

Consider a projectile/plunger p with both static and dynamic friction.
BlasterSim uses an effective friction pressure $p_{\text{f},i}$ defined to oppose the direction of motion.
Using a pressure instead of force allows for more familiar units and is conventional in spud guns.
The maximum static friction pressure is $p_{\text{fs},i}$ and the magnitude of the dynamic friction pressure is $p_{\text{fd},i}$.
When static friction is active, the actual static friction force is just enough force to equilibrate the forces so that p doesn't move (where $\dv{\dot{x}_i}{t} = 0$; see~\eqref{Newton's second law}), until the force exceeds the maximum static friction force.
In other words, the actual static friction pressure $p_{\text{f0},i}$ is the minimum of the equilibrium pressure $p_{\text{fe},i}$ or the maximum static friction pressure, and in the direction of the equilibrium pressure.
Mathematically, this can be stated as
\begin{equation}
    p_{\text{f0},i} = \sgn(p_{\text{fe},i}) \cdot \min(p_{\text{fs},i}, |p_{\text{fe},i}|). \label{eq:actual static friction min}
\end{equation}

The actual dynamic friction pressure is $\sgn(\dot{x}_i) \cdot p_{\text{fd},i}$, taking into account both the magnitude of the dynamic friction pressure $p_{\text{fd},i}$ and the direction.

The effective friction pressure $p_{\text{f},i}$ then will switch between static and dynamic friction depending on the velocity:
\begin{equation}
    p_{\text{f},i} =
    \begin{cases} 
        p_{\text{f0},i},                       & \text{if}~\dot{x}_i = 0~\text{(static)},    \\
        \sgn(\dot{x}_i) \cdot p_{\text{fd},i}, & \text{if}~\dot{x}_i \neq 0~\text{(dynamic)}.
    \end{cases}\label{eq:effective friction pressure cases}
\end{equation}

\textbf{BlasterSim does not use this model} because it is not smooth.
Smooth functions are useful for automatic differentiation (see~\secref{autodiff}) and beneficial for numerical stability.
Also, the use of strict floating point equality in the $\dot{x}_i = 0$ branch can lead to numerical reproducibility issues.
BlasterSim uses a smoothed version of this model which avoids these problems.

To create a smoothed version of~\eqref{effective friction pressure cases}, it's helpful to first rewrite~\eqref{effective friction pressure cases} in terms of the $\sgn$ function:
\begin{equation}
    p_{\text{f},i} =
    p_{\text{f0},i}
    + [p_{\text{fd},i} - \sgn(\dot{x}_i) \cdot p_{\text{f0},i}] \cdot \sgn(\dot{x}_i).
\end{equation}

% See 2025-09-13 handwritten notes for this.
The smooth approximation to~\eqref{effective friction pressure cases} will be made by noting that
\begin{equation}
    \sgn(\dot{x}_i) \approx \tanh\left(\frac{\dot{x}_i}{v_\text{s}}\right),
\end{equation}
where $v_\text{s}$ is a small velocity scale over which the $\tanh$ function changes from -1 to 1.

Consequently, the function implemented in cva.f90 as \texttt{p\_f} is
\begin{equation}
    p_{\text{f},i} =
    p_{\text{f0},i}
    + \left[p_{\text{fd},i} - \tanh\left(\frac{\dot{x}_i}{v_\text{ss}}\right) \cdot p_{\text{f0},i}\right] \cdot \tanh\left(\frac{\dot{x}_i}{v_\text{sd}}\right). \label{eq:p_f}
\end{equation}
In~\eqref{p_f}, instead of a single $v_\text{s}$, two different velocity scales $v_\text{ss}$ and $v_\text{sd}$ are used to allow for finer control.
Note that contrary to the non-smooth model (\eqref{effective friction pressure cases}), the absolute value of~\eqref{p_f} is not guaranteed to be at most $p_{\text{fs},i}$ or $p_{\text{fd},i}$.
To avoid the magnitude of the friction pressure from going above what a user likely intended, $v_\text{ss}$ is made much smaller than $v_\text{sd}$.
The bounds violation comes from both $\tanh$ terms changing at the same time, so it is believed making the two scales very different so that one finishes changing sooner than the other will avoid this problem.

% See 2025-09-30 handwritten notes for part of this.
Making a smooth approximation for $p_{\text{f0},i}$ is more difficult.
Rather than replacing a single part of~\eqref{actual static friction min}, an new function was constructed that does not resemble~\eqref{actual static friction min} algebraically but takes similar values.
A piecewise function is implemented in cva.f90 as \texttt{p\_f0} so that the actual static friction pressure $p_{\text{f0},i}$ equals the equilibrium pressure $p_{\text{fe},i}$ for small values of $p_{\text{fe},i}$:
\begin{equation}
    p_{\text{f0},i} =
    \begin{cases} 
        p_{\text{fe},i},                                         & \text{if}~|p_{\text{fe},i}| \leq p_\text{s}, \\
        \sgn(p_{\text{fe},i}) \cdot p_{\text{f0},i,\text{high}}, & \text{if}~|p_{\text{fe},i}| >    p_\text{s},
    \end{cases}\label{eq:effective friction pressure smooth}
\end{equation}
where $p_\text{s}$ is chosen to be a somewhat arbitrary fraction of $p_{\text{fs},i}$ and
\begin{equation}
    p_{\text{f0},i,\text{high}} = p_{\text{fs},i} \tanh\left(\frac{|p_{\text{fe},i}| - p_\text{s}}{p_{\text{fs},i} - p_\text{s}} + \arctanh\left(\frac{p_\text{s}}{p_{\text{fs},i}}\right)\right).
\end{equation}

Note that despite the appearance of $\sgn(p_{\text{fe},i})$ and $|p_{\text{fe},i}|$,
\eqref{effective friction pressure smooth} % chktex 2
is continuous and has continuous derivatives as those terms only become discontinuous at $p_{\text{fe},i} = 0$ and do not appear in the branch where $p_{\text{fe},i} = 0$.

\subsection{Plunger impact}\label{plunger-impact}

At the moment, BlasterSim does not handle plunger impact with the end of the plunger tube, and will crash if that occurs.
Plunger impact will be handled in a future version of BlasterSim.
Plunger impact does not appear to be necessary in the springers simulated so far, but certainly it is a factor in some springers.

\section{Exterior ballistics}\label{exterior-ballistics}
% TODO

\section{Numerical methods}\label{numerical-methods}

\subsection{Time integration}\label{time-integration}
% TODO

~\cite[p.~1081]{kreyszig_advanced_1993}
~\cite{henon_numerical_1982}

\subsection{Automatic differentiation}\label{autodiff}
% TODO

% justify derivative matching mentioned above, smoothing

~\cite{trettel_btrettelflt_2026}

%\subsection{Uncertainty quantification}\label{uncertainty-quantification}
% TODO
% putko_approach_2001
% TODO: Update `\secref{road-map}` when adding.

%\subsection{Optimization}\label{optimization}
% TODO
% luke_essentials_2013
% Constraints: deb_efficient_2000 (also discuss modification of this)
% TODO: Update `\secref{road-map}` when adding.

% TODO: Also discuss robust optimization
% putko_approach_2001
